---
title: "Usage"
---

```{r, echo = FALSE}
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```

## How Pipes Work

The pipe operator simply passes the output from the upstream function into the first argument of the downstream
function. Hence, these two expressios are equivalent:

```{r}
mean(bfi$age)
bfi$age |> mean()
```

The data piped into a function only fill the first argument. We're free to specify any additional inputs in the normal
way. Hence, the following to expressions are also equivalent.

```{r}
var(bfi$a1, na.rm = TRUE)
bfi$a1 |> var(na.rm = TRUE)
```

### What if the dataset isn’t the first argument?

Many R functions, particularly those that use a so-called *formula interface*, don’t take the input data as their first
argument. If you try to include such a function in a normal pipeline, it won’t work.

```{r, error = TRUE}
bfi |> lm(extra ~ age)
```

In these cases, you can use the special placeholder token `_` to tell R explicitly where to insert the piped object.

```{r}
bfi |> lm(extra ~ age, data = _)
```

Note that we must name the argument to which we assign `_`. The following won't work.

```{r, error = TRUE}
bfi |> lm(extra ~ age, _)
```

This trick makes it possible to use nearly any function in a pipeline, even when the data aren’t the first argument.

```{webr-r}
#| autorun: true
#| context: setup
library(psychTools)
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```

::: practice

Use pipes to perform the following data manipulations on the `bfi` dataset:

1. Create a new variable `age_centered` that centers `age` around its mean.
2. Filter the dataset to include only participants with `age` greater than 25.
3. Select only the new centered age variable, the scores for agreeableness and conscientiousness, and the demographic variables `gen` and `ed`.
4. Display the first 20 rows of the resulting dataset.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

```{r}
bfi |>
  mutate(age_centered = age - mean(age, na.rm = TRUE)) |>
  filter(age > 25) |>
  select(age_centered, agree, consc, gen, ed) |>
  head(20)
```
::::

:::


```{webr-r}
#| autorun: true
#| context: setup
library(psychTools)
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```


::: practice

Use a pipeline to calculate the square root of the mean of the agreeableness scale score for males in the `bfi` data.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.hint >}}

You can use the `unlist()` function to covert a list to a vector.

## {{< var tab_title.solution >}}

```{r}
bfi |>
  filter(gen == "male") |>
  select(agree) |>
  unlist() |>
  mean(na.rm = TRUE) |>
  sqrt()
```
::::

:::
