---
title: "Different Pipe Operators"
---

```{r, echo = FALSE}
library(magrittr)
library(dplyr)
library(psych)

data(bfi)
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")

bfi <- data.frame(bfi, tmp$scores) |>
  mutate(gender = factor(gender, labels = c("male", "female"))) |>
  rename_with(.fn = tolower, .cols = matches("\\d$"))
```

## The **dplyr** Pipe, `%>%`

R's native pipe operator, `|>`, was only introduced in R 4.1. Before that, we used the analagous operator provided by
the **dplyr** package, `%>%`. The **dplyr** pipe is almost exactly the same as the base R version. For example, we can
re-create our first motivating example as follows.

```{r}
library(dplyr)

bfi[c("agree", "consc", "extra", "neuro", "open")] %>%
  cov(use = "pairwise") %>%
  diag() %>%
  sqrt()
```

There are a few minor differences between `|>` and `%>%`. 

1. The **dplyr** pipe uses the `.` character as the dataset placeholder symbol.

```{r}
mtcars %>%
  lm(mpg ~ hp + disp, data = .) %>%
  summary()
```

1. The **dplyr** pipe allows more flexible usage of the dataset placeholder symbol.

```{r}
mtcars %>%
  lm(mpg ~ hp + disp, data = ., weights = .$wt) %>%
  summary()
```

```{r, error = TRUE
mtcars |>
  lm(mpg ~ hp + disp, data = _, weights = _$wt) |>
  summary()
```

## The Exposition Pipe, `%$%`

The exposition pipe, `%$%`, is a very useful flavor or pipe operator provided by the **magrittr** package. The
exposition pipe offers an elegent alternative to the dataset placeholder symbol in many common types of pipeline. Most
notably, pipelines involving functions that employ the so-called *formula interface*, such as `lm()`, `t.test()`, or
`boxplot()`.

Instead of passing a the processed dataset into the first argument of a downstream function, the exposition pipe
*exposes* the column names of the processed data frame to the downstream function. For example, the following code uses
the exposition pipe to regress `age` on `extra` in the `bfi` data.

```{r}
bfi %$% lm(extra ~ age)
```

The exposition pipe allows the `lm()` call to access the colum names of `bfi` when resovling the model formula `extra ~
age`. We can achieve analagous results in several ways.

```{r}
## Explicitly regress two vectors:
lm(bfi$extra ~ bfi$age)

## Use the Base R equivelant of %$%:
with(bfi, lm(extra ~ age))

## Use a dataset placeholder symbol:
bfi |> lm(extra ~ age, data = _)
```

## The Assignment Pipe, `%<>%`

The final type of pipe that we'll cover is certainly a niche tool, but it provides a nice little bit of syntatic sugar.
We only care about assignment pipes in particular types of pipelines. Names, pipelines with two characteristics:

1. The pipeline modifies a dataset
1. The pipeline overwrites the original dataset with the modified version

For example, the following pipeline uses **dplyr** data manipulation functions to modify the `bfi1` dataset and
overwrite the original with the modified version produced by the pipeline.

```{r}
bfi1 <- bfi
head(bfi1)

bfi1 <- bfi1 %>%
  filter(gender == "female") %>%
  select(-matches("\\d$"))

head(bfi1)
```

The assignment pipe cleans up the syntax a little but and makes the destructive nature of the pipeline clear. When you
see an assignment pipe, you know that the modified data produced by the pipeline will replace the original dataset.

```{r}
bfi1 <- bfi
head(bfi1)

bfi1 %<>%
  filter(gender == "female") %>%
  select(-matches("\\d$"))

head(bfi1)
```

Note that we must use the **dplyr** pipe, `%>%`, if we include an assignment pipe in our pipeline. The Base R pipe
doesn't play nicely with the assignment pipe. More specifically, the assignment pipe behaves like a basic pipe (i.e.,
`|>` or `%>%`) when combined with the Base R pipe operator, as you can see below.

```{r}
bfi1 <- bfi
head(bfi1)

bfi1 %<>%
  filter(gender == "female") |>
  select(-matches("\\d$"))

head(bfi1)
```

```{webr-r}
#| autorun: true
#| context: setup
library(psychTools)
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```


::: practice

## Practice

Use the pipe and exposition pipe to calculate the correlation between `age` and `agree` for adults in the `bfi` data.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.hint >}}

You can use the `cor()` function to compute the correlation between two variables.

## {{< var tab_title.solution >}}

```{r}
bfi %>%
  filter(age >= 18) %$%
  cor(age, agree)
```
::::

:::

